---
title: "EDA.Rmd"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read in the data.
```{r}
df = read.csv("data.csv")
```
Check variables for type issues etc.
```{r}
str(df)
```
Let's Return.to.Prison and Target.Population boolean
```{r}
df$Return.to.Prison = as.character(df$Return.to.Prison)
df$Return.to.Prison[df$Return.to.Prison == "Yes"] = "TRUE"
df$Return.to.Prison[df$Return.to.Prison == "No"] = "FALSE"
df$Return.to.Prison = as.logical(df$Return.to.Prison)

df$Target.Population = as.character(df$Target.Population)
df$Target.Population[df$Target.Population == "Yes"] = "TRUE"
df$Target.Population[df$Target.Population == "No"] = "FALSE"
df$Target.Population = as.logical(df$Target.Population)

str(df)
```

Let's get rid of NAs.
```{r}
# find all columns with NA using an apply of anyNA on each column
colnames(df)[apply(df, 2, anyNA)]
```
Just Days.to.Return it seems. These are presumably those who did not reoffend, so these are valid values, so we will leave it as is. Let's see how severe it is.
```{r}
cat("Total NA rows:", sum(is.na(df$Days.to.Return)),"\n")
cat("Rows total:", nrow(df), "\n")
cat("Proportion of rows NA:", sum(is.na(df$Days.to.Return)) / nrow(df))
```
So around 2/3 of released prisoners did not return to prison, assuming NAs are not just missing values. Given the cleanliness of the dataset, this likely is a good assumption. We can check by seeing if these NAs are the same rows as FALSE values for Return.to.Prison:
```{r}
# we can see if there any NAs that aren't matched with those that did not return to prison
length(df$Days.to.Return[(is.na(df$Days.to.Return)) & (df$Return.to.Prison == TRUE)])
```
Since there are no instances of a return to prison with an NA days to return, we can presume all of the NAs are valid.



Now we can look at the values of some of our most relevant factor variables.
```{r}
levels(df$Offense.Classification)
```
```{r}
levels(df$Offense.Type)
```
```{r}
levels(df$Offense.Subtype)
```
We will likley use dummy variables if we use these for regression, and we likely will considering their potential usefulness in predicting recidivism, new offense, days until return, etc.

Let's see if offense types and subtypes differ between new and original offenses.
```{r}
levels(df$New.Offense.Type)
```
So New.Offense.Type adds Assault, Drug Possession, Flight/Escape, Traffic, and Sex. These are all categories previously contained in offense subtype.

```{r}
levels(df$New.Offense.Sub.Type)
```
The new offense type categories are still within subtype, though there are some differences between the subtype factors. Now, "Special sentence Revocation" is gone and we have a new blank factor, for example. Also note that new offense type has a blank factor. Maybe these are markers that the new offense matched the old offense type and/or subtype. Let's see if old and new types match otherwise:
```{r}
sum(as.character(df$Offense.Type) == as.character(df$New.Offense.Type))
```
So, we have around 4000 indices where offense type and new offensive type match. Let's see what the values are. It is then unclear what the blank factors mean. 
```{r}
length(df$New.Offense.Type[(df$New.Offense.Type == "") | (df$New.Offense.Sub.Type == "")])
```
So, these occur very often in the data set. It is still possible that these indicate a match of original and new types, but it may warrent further investigation. They are often likely released prisoners who never reoffended, similar to NAs in Days.to.Return.

Let's reorder variables by frequency and plot their distributions.
```{r}
library(ggplot2)

df$Offense.Type = reorder(df$Offense.Type, df$Offense.Type, length)
df$Offense.Subtype = reorder(df$Offense.Subtype, df$Offense.Subtype, length)
df$New.Offense.Type = reorder(df$New.Offense.Type, df$New.Offense.Type, length)
df$New.Offense.Sub.Type = reorder(df$New.Offense.Sub.Type, df$New.Offense.Sub.Type, length)
df$Age.At.Release = reorder(df$Age.At.Release, df$Age.At.Release, length)
```
Plots:
```{r}
ggplot(df, aes(x=Offense.Type)) + geom_bar() + labs(x="Offense Type", y="Count", title="Count of Offenses by Type") + coord_flip()
```
So, drug and proprtyb charges are our most common first offenses.
```{r}
ggplot(df, aes(x=Offense.Subtype)) + geom_bar() + labs(x="Offense Subtype", y="Count", title="Count of Offenses by Subtype") + coord_flip()
```
Trafficking is very high relative to its nearest competitors, which are assault and burglary.
```{r}
ggplot(df, aes(x=New.Offense.Type)) + geom_bar() + labs(x="New Offense Type", y="Count", title="Count of New Offenses by Type") + coord_flip()
```
We can see the vast majority of new offense types are empty values. Since these are so common, it is likely a mixture of non-reoffenders and those who reoffended with the same offense type. Besides empty values, drug and property offenses are most common.

What about subtypes?
```{r}
ggplot(df, aes(x=New.Offense.Sub.Type)) + geom_bar() + labs(x="New Offense Subtype", y="Count", title="Count of New Offenses by Subtype") + coord_flip()
```
Again, the empty factor is by far the most common, followed by trafficking, theft, then assault.

Now, let's look at distribuitions for Sex, Day.to.Return, and Age.at.Release.
```{r}
ggplot(df, aes(x=Sex)) + geom_bar() + labs(x="Sex", y="Count", title="Count of Offenders by Sex") + coord_flip()
```
Mostly male.

```{r}
days.mean = mean(df$Days.to.Return, na.rm=TRUE)
ggplot(df, aes(x=Days.to.Return)) + geom_histogram() + labs(x="Days to Return", y="Count", title="Distribution of Offenders by Days to Return") + geom_vline(xintercept = days.mean,colour="blue")
```
It looks like we have a right skewed distribution of days until return to prison with a mean of ~470.

```{r}
ggplot(df, aes(x=Age.At.Release)) + geom_bar() + labs(x="Age at Release", y="Count", title="Count of Offenders by Age Group at Release")
```
```
So, most offenders are 25-34 with the next highest group being 35-44.


