---
title: "final_project_reg_question"
author: "Max Ruehle"
date: "October 25, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read in the data.
```{r}
df = read.csv("data_cleaned.csv")
```


# Days Until Reinstatement (Regression)

One big question that we hope to be answered by this dataset is, given a prisoner eventually returns to prison, how many days will it be until they return?

Let's first look at how many prisoners actually returned to prison in this three-year span:

```{r}
barplot(table(df$Return.to.Prison), main = "Return to Prison?", names.arg = c("No", "Yes"))

library(dplyr)

count(df, Return.to.Prison)
```

We can see that of the 26,020 prisoners released, 8,681 ended up returning to prison within the three-year span.

Since we only want to look at those that returned, we can now subset the data:

```{r}
df = filter(df, Return.to.Prison == TRUE)
count(df)
```

Now, we can subset the columns that are important for building our regression model and because they are all categorical data, we can turn them into factor variables for extra insurance:

Note: I do believe that Offense.Subtype is important in the model, but the sake of it having 26 different factor levels, I have left it out for now.

```{r}
df = select(df, Release.Type, Race...Ethnicity, Age.At.Release, Sex, Offense.Classification, Offense.Type, Offense.Subtype, Recidivism.Type, Days.to.Return)

df$Release.Type = factor(df$Release.Type)
df$Race...Ethnicity = factor(df$Race...Ethnicity)
df$Age.At.Release = factor(df$Age.At.Release)
df$Sex = factor(df$Sex)
df$Offense.Classification = factor(df$Offense.Classification)
df$Offense.Type = factor(df$Offense.Type)
df$Offense.Subtype = factor(df$Offense.Subtype)
df$Recidivism.Type = factor(df$Recidivism.Type)
```


First, let's take a very simple model using just the Sex of the offenders. This will give us an idea of how our models will work.
```{r}
contrasts(df$Sex)

summary(lm(Days.to.Return~Sex, data = df))
```

This isn't too difficult to understand. Our model simply says if the offender is female, the value for SexMale will be 0, giving us a Days.to.Return value of approximately 457 days. Compare this to if the offender is Male, the Days.to.Return value would be about 471 days.

Now, we obviously cannot use this model for many reasons, one of them being the p-value we get from our chosen factor is 0.131, which is quite high.


Let's build a new model using Sex, Offense Type, and Recidivism Type:

```{r}
contrasts(df$Sex)
contrasts(df$Offense.Type)
contrasts(df$Recidivism.Type)

summary(lm(Days.to.Return~Sex + Offense.Type + Recidivism.Type, data = df))
```

We can see that this model is quite an improvement over the last one. Let's split our data to get our test MSE.

```{r}
n = dim(df)[1]
train_index = sample(1:n,7000,rep=FALSE)
train_df = df[train_index,]
test_df = df[-train_index,]

fit = lm(Days.to.Return~Sex + Offense.Type + Recidivism.Type, data = train_df)

X_test = test_df[,-9]
predicted_values = predict(fit, X_test)
sum((test_df$Days.to.Return - predicted_values)^2)/(n/2)
```

We would like to get a better model than this. The reason for doing so is that we essentially just chose three variables that we thought could be good. But now, by using Stepwise selection, we can create a much better model that is more efficient and only keeps in the factors that go into determining recidivism.

```{r}
model0 = lm(Days.to.Return~1,data=df)
summary(model0)

modelfull = lm(Days.to.Return~.,data=df)
summary(modelfull)

step(model0,scope=list(lower=model0,upper=modelfull),direction="forward")
step(modelfull,scope=list(lower=model0,upper=modelfull),direction="backward")
step(modelfull,scope=list(lower=model0,upper=modelfull),direction="both")
```

Using forward, backward, and hybrid stepwise selection gives us a model with all factors included. This is due to there not being enough factors in our set. In each forward step, the Test MSE would decrease and in each backward step, the Test MSE would increase. This tells us that using all factors would give us the lowest Test MSE.

```{r}
predicted_values = predict(modelfull, X_test)
sum((test_df$Days.to.Return - predicted_values)^2)/(n/2)
```

It is obvious that much more data is necessary to accurately produce the time it takes to predict when an inmate will commit another crime. This is a great example of how, with more data, we could come closer to predicting when when someone will commit another crime, but we also have to keep in mind that data cannot tell us the full story. It is extremely hard to predict what a person will do because there is often different motives for why someone needs to do something (e.g. commiting a crime).



