---
title: "final_project_reg_question"
author: "Max Ruehle"
date: "October 25, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read in the data.
```{r}
df = read.csv("data_cleaned.csv")
```


# Days Until Reinstatement (Regression)

One big question that we hope to be answered by this dataset is, given a prisoner eventually returns to prison, how many days will it be until they return?

Let's first look at how many prisoners actually returned to prison in this three-year span:

```{r}
barplot(table(df$Return.to.Prison), main = "Return to Prison?", names.arg = c("No", "Yes"))

library(dplyr)

count(df, Return.to.Prison)
```

We can see that of the 26,020 prisoners released, 8,681 ended up returning to prison within the three-year span.

Since we only want to look at those that returned, we can now subset the data:

```{r}
df = filter(df, Return.to.Prison == TRUE)
count(df)
```

Now, we can subset the columns that are important for building our regression model and because they are all categorical data, we can turn them into factor variables for extra insurance:

Note: I do believe that Offense.Subtype is important in the model, but the sake of it having 26 different factor levels, I have left it out for now.

```{r}
df = select(df, Release.Type, Race...Ethnicity, Age.At.Release, Sex, Offense.Classification, Offense.Type,  Recidivism.Type, Days.to.Return)

df$Release.Type = factor(df$Release.Type)
df$Race...Ethnicity = factor(df$Race...Ethnicity)
df$Age.At.Release = factor(df$Age.At.Release)
df$Sex = factor(df$Sex)
df$Offense.Classification = factor(df$Offense.Classification)
df$Offense.Type = factor(df$Offense.Type)
df$Recidivism.Type = factor(df$Recidivism.Type)
```


First, let's take a very simple model using just the Sex of the offenders. This will give us an idea of how our models will work.
```{r}
contrasts(df$Sex)

summary(lm(Days.to.Return~Sex, data = df))
```

This isn't too difficult to understand. Our model simply says if the offender is female, the value for SexMale will be 0, giving us a Days.to.Return value of approximately 457 days. Compare this to if the offender is Male, the Days.to.Return value would be about 471 days.

Now, we obviously cannot use this model for many reasons, one of them being the p-value we get from our chosen factor is 0.131, which is quite high.


Let's build a new model using Sex, Offense Type, and Recidivism Type:

```{r}
contrasts(df$Sex)
contrasts(df$Offense.Type)
contrasts(df$Recidivism.Type)

summary(lm(Days.to.Return~Sex + Offense.Type + Recidivism.Type, data = df))
```

We can see that this model is quite an improvement over the last one. Let's split our data to get our test MSE.

```{r}
n = dim(df)[1]
train_index = sample(1:n,n/2,rep=FALSE)
train_df = df[train_index,]
test_df = df[-train_index,]

fit = lm(Days.to.Return~Sex + Offense.Type + Recidivism.Type, data = train_df)

X_test = test_df[,-8]
predicted_values = predict(fit, X_test)
sum((test_df$Days.to.Return - predicted_values)^2)/(n/2)
```

This test MSE is obviously very high. Continuing, I will use stepwise selection to fine tune my model.




